\replaced{Fast Fourier transformation of electromagnetic data for expensive kernels}{Fast time-domain electromagnetic modelling in the frequency domain}

### 2020-11-27 ManuscriptTime.tex
EOD

{\color{alizarin}
\verb!v v v REMOVE COMMENTS v v v REMOVE COMMENTS v v v REMOVE COMMENTS v v v!
}
{\color{mygreen}
\begin{itemize}
  \item Too generic; more references
  \item what are the methodological advantages with respect to other
    approaches.
\end{itemize}

@@

The first example is the inline electric field from a source at the origin
measured by an inline receiver with an offset of 900\,m in a homogeneous space
of 1\,\ohmm. We chose this simple example to compare it with the analytical
solution and with previously published results. We used the following values to
define the required frequencies: $f_\mr{min}=0.05\,$Hz, $f_\mr{max}=21\,$Hz,
using FFTLog with 5 frequencies per decade. This results in 14 frequencies to
compute from 0.05\,Hz to 20.0\,Hz. The complete frequency range for the
transform includes 30 frequencies from 0.0002\,Hz to 126.4\,Hz. For the
adaptive gridding the following inputs were used: $n_\delta=12$, minimum cell
width must be between 20 and 40\,m, and $\alpha_\mr{s}=1$,
$\alpha_\mr{c}=[1,1.3]$. This created grids with cell numbers between
\num{46080} ($80\times24\times24$, for 20.0\,Hz) and \num{128000}
($80\times40\times40$, for 0.05\,Hz) cells. The run times for each frequency,
the corresponding number of cells, minimum cell width and computation domain
stretching factor are listed in Table~\ref{tbl:timefull}. The total run time to
compute this model was less than two minutes.
%
\begin{table}
\begin{minipage}{\twidth}
  \centering
  \caption{Run times per frequency for the homogeneous space example, with the
    corresponding number of cells and minimum cell width as well as the
    stretching factor in the computation domain; $\alpha_\mr{s}=1$ everywhere.}
  \label{tbl:timefull}
  \begin{tabular}{S[table-format=2.4]rrrr}
    \toprule
    %
    {Frequency} & Time & nx$\times$ny$\times$nz & $\Delta_\mr{min}$ & $\alpha_\mr{c}$ \\
    {(Hz)}  & (s) &  & (m) & \\
    \midrule
    %
    20.0    &\replaced{ 4}{ 3} & 80$\times$24$\times$24 & 20 & 1.26 \\
    12.6    &\replaced{ 8}{ 6} & 96$\times$32$\times$32 & 20 & 1.17 \\
     7.98   &\replaced{ 8}{ 8} & 96$\times$32$\times$32 & 20 & 1.20 \\
     5.03   &\replaced{ 8}{ 7} & 96$\times$32$\times$32 & 20 & 1.23 \\
     3.18   &\replaced{ 7}{ 7} & 80$\times$32$\times$32 & 24 & 1.23 \\
     2.00   &\replaced{ 7}{ 5} & 80$\times$32$\times$32 & 30 & 1.21 \\
     1.26   &\replaced{ 5}{ 4} & 64$\times$32$\times$32 & 37 & 1.21 \\
     0.798  &\replaced{ 5}{ 5} & 64$\times$32$\times$32 & 40 & 1.23 \\
     0.503  &\replaced{ 5}{ 5} & 64$\times$32$\times$32 & 40 & 1.26 \\
     0.318  &\replaced{ 5}{ 5} & 64$\times$32$\times$32 & 40 & 1.28 \\
     0.200  &\replaced{ 8}{10} & 64$\times$40$\times$40 & 40 & 1.27 \\
     0.126  &\replaced{ 8}{10} & 64$\times$40$\times$40 & 40 & 1.30 \\
     0.0798 &\replaced{10}{14} & 80$\times$40$\times$40 & 40 & 1.26 \\
     0.0503 &\replaced{11}{15} & 80$\times$40$\times$40 & 40 & 1.28 \\
    %
    \bottomrule
  \end{tabular}
\end{minipage}
\end{table}
%

@@

The second example is a shallow marine, layered model with 200\,m of seawater
(3\,S/m) above a halfspace of 1\,S/m, an embedded target layer at 2\,km depth,
100\,m thick, with a conductivity of 0.02\,S/m. The source is located 20\,m
above the seafloor and the receivers are on the seafloor. We chose the
frequency range such that we can model offsets from 3 to 7 kilometers, with
$f_\mr{min}=0.007\,$Hz and $f_\mr{max}=32\,$Hz, using FFTLog with 5 frequencies
per decade. This results in computations for 19 frequencies from 0.008\,Hz to
31.8\,Hz. The complete frequency range for the transform includes 35
frequencies from $2\,10^{-5}$ to $126.4\,$Hz. For the adaptive gridding, we
used a cell width of 100\,m in the core domain and stretching outside up to a
factor 1.5, where the computation domain extends up to 50\,km in each
direction. This yielded grids between \num{204800} (higher frequencies) and
\num{245760} (lower frequencies) cells. The run times for each frequency and
their corresponding parameters are listed in Table~\ref{tbl:timemarine}.
%
\begin{table}
\begin{minipage}{\twidth}
  \centering
  \caption{Run times per frequency for the marine 1D example, with the
    corresponding number of cells and minimum cell width as well as the
    stretching factor in the computation domain; $\alpha_\mr{s}=1$ everywhere.}
  \label{tbl:timemarine}
  \begin{tabular}{S[table-format=2.5]rrrr}
    \toprule
    %
    {Frequency} & time & nx$\times$ny$\times$nz & $\Delta_\mr{min}$ & $\alpha_\mr{c}$ \\
    {(Hz)}  & (s) &  & (m) & \\
    \midrule
    %
    31.8     & \replaced{16}{10} & 128$\times$40$\times$40 & 100 & 1.36 \\
    20.0     & \replaced{17}{ 9} & 128$\times$40$\times$40 & 100 & 1.36 \\
    12.6     & \replaced{17}{ 9} & 128$\times$40$\times$40 & 100 & 1.36 \\
     7.98    & \replaced{17}{17} & 128$\times$40$\times$40 & 100 & 1.36 \\
     5.03    & \replaced{17}{21} & 128$\times$40$\times$40 & 100 & 1.44 \\
     3.18    & \replaced{17}{20} & 128$\times$40$\times$40 & 100 & 1.48 \\
     2.00    & \replaced{17}{18} & 128$\times$40$\times$40 & 100 & 1.49 \\
     1.26    & \replaced{20}{22} & 128$\times$40$\times$48 & 100 & 1.36 \\
     0.798   & \replaced{24}{26} & 128$\times$40$\times$48 & 100 & 1.36 \\
     0.503   & \replaced{27}{29} & 128$\times$40$\times$48 & 100 & 1.36 \\
     0.318   & \replaced{27}{42} & 128$\times$40$\times$48 & 100 & 1.36 \\
     0.200   & \replaced{34}{41} & 128$\times$40$\times$48 & 100 & 1.38 \\
     0.126   & \replaced{34}{48} & 128$\times$40$\times$48 & 100 & 1.40 \\
     0.0798  & \replaced{37}{54} & 128$\times$40$\times$48 & 100 & 1.41 \\
     0.0503  & \replaced{37}{59} & 128$\times$40$\times$48 & 100 & 1.44 \\
     0.0318  & \replaced{44}{74} & 128$\times$40$\times$48 & 100 & 1.44 \\
     0.0200  & \replaced{44}{81} & 128$\times$40$\times$48 & 100 & 1.47 \\
     0.0126  & \replaced{37}{96} & 128$\times$40$\times$48 & 100 & 1.48 \\
     0.00798 & \replaced{43}{93} & 128$\times$40$\times$48 & 100 & 1.49 \\
    %
    \bottomrule
  \end{tabular}
\end{minipage}
\end{table}
%

@@

The results are shown in Figure~\ref{fig:3d-result}: In (a) the 1D background
responses and the relative error using the semi-analytical
result\deleted{obtained from \empymod}, and in (b) the responses including the
target.
%
\begin{figure*}
  \centering
  \includegraphics[width=\fwidth]{10-3d-result}
  \caption{Responses for the model outlined in Figure~\ref{fig:3d-model} using
    \replaced{time-domain}{\simpeg} and \replaced{frequency-domain
    computations}{\emg3d}, and for the layered background also \replaced{the
    semi-analytical result}{\empymod}. In the lower plot of (a) the relative
    error (\%) \added{is shown} in comparison to \replaced{the semi-analytical
    result}{\empymod is shown}, and in (b) the normalized difference (\%)
    between the two 3D codes. The results of these plots are clipped for values
    smaller than 0.1\,\% and bigger than 10\,\%.}
  \label{fig:3d-result}
\end{figure*}
%
The background comparison shows that both 3D codes do an acceptable job having
a relative error of a few percents at most; the result \replaced{obtained
through transformation}{for \emg3d} seems to be better at early times. We
cannot compare the errors for the response that includes the target for lack of
an analytical solution. The \added{semi-analytical} background model
\deleted{from \empymod }is only included to show that there is a significant
response from the target. We therefore show the normalized difference
(\replaced{NRMSD}{ND}) between the two \replaced{responses $R_1$ and $R_2$}{3D
codes} as a percentage, where \replaced{$\mr{NRMSD~(\%)} = 200|R_1 -
R_2|/(|R_1| + |R_2|)$.}{$\mr{ND}= 50|(\simpeg-\emg3d)/(\simpeg+\emg3d)|$.} The
\replaced{NRMSD}{ND} between the two codes is below 1\,\% everywhere except for
early times.

\added{¡TODO CHANGE THIS LAST PARAGRAPH!}
Both codes took a little over 7 minutes to compute the two models (single
thread). However, in this particular comparison, the main difference in runtime
is not frequency-domain computation vs. time-domain computation, but iterative
solver vs. direct solver. For a test we doubled the number of cells in each
direction by using 50\,m for the minimum cell width instead of 100\,m, and
taking also half of the stretching, which results then in a mesh of
\num{466752} cells \replaced{in the time domain}{for \simpeg} and meshes
between \num{102400} and \num{614400} \replaced{in the frequency domain}{for
\emg3d}. Because the memory of our test machine was not large enough for the
direct solver of \simpeg, the last model had to be run on a cluster. In this
scenario, \emg3d was about eight times faster (single thread).

>>>>

{\color{alizarin}
\verb!v v v REMOVE COMMENTS v v v REMOVE COMMENTS v v v REMOVE COMMENTS v v v!
}
{\color{mygreen}
\begin{itemize}
  \item Too generic; more references
  \item what are the methodological advantages with respect to other
    approaches.
  \item wants: TEM can be accelerated using similar coarse-gridding
    considerations as mentioned in your manuscript. In Commer \& Newman (2005),
    the FDTD method could be accelerated through grid-coarsening during
    runtime, using the fact that the transient field-decay filters out
    high-frequency components.
  \item I do believe the subspace-projection method should definitely be
    discussed in a pro/con manner.
  \item \ldots
\end{itemize}

Dieter: One one hand we agree that the introduction could be expanded, on the
other hand we are not fond of repeating everything again. The introduction
contains quite some extensive, good review papers (Oristaglio and Spies, 1999;
Avdeev, 2005; Börner, 2010; Constable, 2010). The comparison between different
approaches for time-domain computation was done in quite some detail in Mulder
(2008)

TODO: We think that there is a misunderstanding here. First we mention Avdeev
(2005) and Börner (2010), and say that these are very good review papers up to
2005 and 2010, respectively. Recent publications then means publications since
these two review papers were published. We do not want to repeat again all the
pioneering publications, that has been done, but highlight a few recent ones.
We changed the wording in the introduction to make this clear.

@@

The first example is the inline electric field from a source at the origin
measured by an inline receiver with an offset of 900\,m in a homogeneous space
of 1\,\ohmm. We chose this simple example to compare it with the analytical
solution and with previously published results. We used the following values to
define the required frequencies: $f_\mr{min}=0.05\,$Hz, $f_\mr{max}=21\,$Hz,
using FFTLog with 5 frequencies per decade. This results in 14 frequencies to
compute from 0.05\,Hz to 20.0\,Hz. The complete frequency range for the
transform includes 30 frequencies from 0.0002\,Hz to 126.4\,Hz. For the
adaptive gridding the following inputs were used: $n_\delta=12$, minimum cell
width must be between 20 and 40\,m, and $\alpha_\mr{s}=1$,
$\alpha_\mr{c}=[1,1.3]$. This created grids with cell numbers between
\num{46080} ($80\times24\times24$, for 20.0\,Hz) and \num{128000}
($80\times40\times40$, for 0.05\,Hz) cells. The run times for each frequency,
the corresponding number of cells, minimum cell width and computation domain
stretching factor are listed in Table~\ref{tbl:timefull}. The total run time to
compute this model was less than two minutes.
%
\begin{table}
\begin{minipage}{\twidth}
  \centering
  \caption{Run times per frequency for the homogeneous space example, with the
    corresponding number of cells and minimum cell width as well as the
    stretching factor in the computation domain; $\alpha_\mr{s}=1$ everywhere.}
  \label{tbl:timefull}
  \begin{tabular}{S[table-format=2.4]rrrr}
    \toprule
    %
    {Frequency} & Time & nx$\times$ny$\times$nz & $\Delta_\mr{min}$ & $\alpha_\mr{c}$ \\
    {(Hz)}  & (s) &  & (m) & \\
    \midrule
    %
    20.0    &  4 & 80$\times$24$\times$24 & 20 & 1.26 \\
    12.6    &  8 & 96$\times$32$\times$32 & 20 & 1.17 \\
     7.98   &  8 & 96$\times$32$\times$32 & 20 & 1.20 \\
     5.03   &  8 & 96$\times$32$\times$32 & 20 & 1.23 \\
     3.18   &  7 & 80$\times$32$\times$32 & 24 & 1.23 \\
     2.00   &  7 & 80$\times$32$\times$32 & 30 & 1.21 \\
     1.26   &  5 & 64$\times$32$\times$32 & 37 & 1.21 \\
     0.798  &  5 & 64$\times$32$\times$32 & 40 & 1.23 \\
     0.503  &  5 & 64$\times$32$\times$32 & 40 & 1.26 \\
     0.318  &  5 & 64$\times$32$\times$32 & 40 & 1.28 \\
     0.200  &  8 & 64$\times$40$\times$40 & 40 & 1.27 \\
     0.126  &  8 & 64$\times$40$\times$40 & 40 & 1.30 \\
     0.0798 & 10 & 80$\times$40$\times$40 & 40 & 1.26 \\
     0.0503 & 11 & 80$\times$40$\times$40 & 40 & 1.28 \\
    %
    \bottomrule
  \end{tabular}
\end{minipage}
\end{table}
%

@@

The second example is a shallow marine, layered model with 200\,m of seawater
(3\,S/m) above a halfspace of 1\,S/m, an embedded target layer at 2\,km depth,
100\,m thick, with a conductivity of 0.02\,S/m. The source is located 20\,m
above the seafloor and the receivers are on the seafloor. We chose the
frequency range such that we can model offsets from 3 to 7 kilometers, with
$f_\mr{min}=0.007\,$Hz and $f_\mr{max}=32\,$Hz, using FFTLog with 5 frequencies
per decade. This results in computations for 19 frequencies from 0.008\,Hz to
31.8\,Hz. The complete frequency range for the transform includes 35
frequencies from $2\,10^{-5}$ to $126.4\,$Hz. For the adaptive gridding, we
used a cell width of 100\,m in the core domain and stretching outside up to a
factor 1.5, where the computation domain extends up to 50\,km in each
direction. This yielded grids between \num{204800} (higher frequencies) and
\num{245760} (lower frequencies) cells. The run times for each frequency and
their corresponding parameters are listed in Table~\ref{tbl:timemarine}.
%
\begin{table}
\begin{minipage}{\twidth}
  \centering
  \caption{Run times per frequency for the marine 1D example, with the
    corresponding number of cells and minimum cell width as well as the
    stretching factor in the computation domain; $\alpha_\mr{s}=1$ everywhere.}
  \label{tbl:timemarine}
  \begin{tabular}{S[table-format=2.5]rrrr}
    \toprule
    %
    {Frequency} & time & nx$\times$ny$\times$nz & $\Delta_\mr{min}$ & $\alpha_\mr{c}$ \\
    {(Hz)}  & (s) &  & (m) & \\
    \midrule
    %
    31.8     & 16 & 128$\times$40$\times$40 & 100 & 1.36 \\
    20.0     & 17 & 128$\times$40$\times$40 & 100 & 1.36 \\
    12.6     & 17 & 128$\times$40$\times$40 & 100 & 1.36 \\
     7.98    & 17 & 128$\times$40$\times$40 & 100 & 1.36 \\
     5.03    & 17 & 128$\times$40$\times$40 & 100 & 1.44 \\
     3.18    & 17 & 128$\times$40$\times$40 & 100 & 1.48 \\
     2.00    & 17 & 128$\times$40$\times$40 & 100 & 1.49 \\
     1.26    & 20 & 128$\times$40$\times$48 & 100 & 1.36 \\
     0.798   & 24 & 128$\times$40$\times$48 & 100 & 1.36 \\
     0.503   & 27 & 128$\times$40$\times$48 & 100 & 1.36 \\
     0.318   & 27 & 128$\times$40$\times$48 & 100 & 1.36 \\
     0.200   & 34 & 128$\times$40$\times$48 & 100 & 1.38 \\
     0.126   & 34 & 128$\times$40$\times$48 & 100 & 1.40 \\
     0.0798  & 37 & 128$\times$40$\times$48 & 100 & 1.41 \\
     0.0503  & 37 & 128$\times$40$\times$48 & 100 & 1.44 \\
     0.0318  & 44 & 128$\times$40$\times$48 & 100 & 1.44 \\
     0.0200  & 44 & 128$\times$40$\times$48 & 100 & 1.47 \\
     0.0126  & 37 & 128$\times$40$\times$48 & 100 & 1.48 \\
     0.00798 & 43 & 128$\times$40$\times$48 & 100 & 1.49 \\
    %
    \bottomrule
  \end{tabular}
\end{minipage}
\end{table}
%

@@

The results are shown in Figure~\ref{fig:3d-result}: In (a) the 1D background
responses and the relative error using the semi-analytical
result\deleted{obtained from \empymod}, and in (b) the responses including the
target.
%
\begin{figure*}
  \centering
  \includegraphics[width=\fwidth]{10-3d-result}
  \caption{Responses for the model outlined in Figure~\ref{fig:3d-model} using
    \replaced{time-domain}{\simpeg} and \replaced{frequency-domain
    computations}{\emg3d}, and for the layered background also \replaced{the
    semi-analytical result}{\empymod}. In the lower plot of (a) the relative
    error (\%) \added{is shown} in comparison to \replaced{the semi-analytical
    result}{\empymod is shown}, and in (b) the normalized difference (\%)
    between the two 3D codes. The results of these plots are clipped for values
    smaller than 0.1\,\% and bigger than 10\,\%.}
  \label{fig:3d-result}
\end{figure*}
%
The background comparison shows that both 3D codes do an acceptable job having
a relative error of a few percents at most; the result \replaced{obtained
through transformation}{for \emg3d} seems to be better at early times.
\added{The reason is probably the implemented backward Euler in the time-domain
code which has an error of order one in time.} We cannot compare the errors for
the response that includes the target for lack of an analytical solution. The
\added{1D} background model \deleted{from \empymod }is only included to show
that there is a significant response from the target. We therefore show the
normalized difference (\replaced{NRMSD}{ND}) between the two
\replaced{responses $R_1$ and $R_2$}{3D codes} as a percentage, where
\replaced{$\mr{NRMSD~(\%)} = 200|R_1 - R_2|/(|R_1| + |R_2|)$.}{$\mr{ND}=
50|(\simpeg-\emg3d)/(\simpeg+\emg3d)|$.} The \replaced{NRMSD}{ND} between the
two codes is below 1\,\% everywhere except for early times. Both codes took
\replaced{roughly 4--5}{a little over 7} minutes to compute the two models
(single thread). However, in this particular comparison, the main difference in
runtime is not frequency-domain computation vs. time-domain computation, but
iterative solver vs. direct solver.

\deleted{\emph{(Deleted last three sentences, as they do not add a lot.)}For a
test we doubled the number of cells in each direction by using 50\,m for the
minimum cell width instead of 100\,m, and taking also half of the stretching,
which results then in a mesh of \num{466752} cells for \simpeg and meshes
between \num{102400} and \num{614400} for \emg3d. Because the memory of our
test machine was not large enough for the direct solver of \simpeg, the last
model had to be run on a cluster. In this scenario, \emg3d was about eight
times faster (single thread).}
